Joint maximum-likelihood of phylogenies and ancestral states is not consistent

### 2018-04-30 appendix.tex
clarify, be explicit, address (most of) reviewer 2

Since we have a finite character alphabet, for a given column $i$ there are a finite number of possible assignments of characters to tips $\alignmentColumn_i$ or internal nodes $\ancestralStateColumn_i$; this results in a simplification of likelihood calculation.
Take the tip labels of $\tau$ to be $\{1,\ldots,\nSiteRows\}$.
For likelihood calculation under the binary symmetric model, we describe a given $\alignmentColumn_i$ as a subset of indices $\siteSplit\subseteq\siteSplitSet:=\{1,\ldots,\nSiteRows-1\}$ with equivalent characters, commonly called a ``site split.''
We define the site split $\siteSplit$ for a $\alignmentColumn_i$ as simply $\alignmentColumn_i$ if the label $\nSiteRows$ is not in $\alignmentColumn_i$, and as its complement otherwise.
Taking such a complement simplifies but does not change the result of likelihood computation because the probability of observing a particular collection of binary characters is equivalent to the probability of its complement under the binary symmetric model.

@@

We first fix notation.
\begin{definition}
Let the mapping from site patterns to site splits be
$$
\patternToSplit:\alphabet^\nSiteRows\rightarrow\mathcal{P}(\siteSplitSet)
$$
and the mapping from ancestral states and tip states to ancestral state splits be
$$
\ancestralToSplit:\alphabet^\nAncestralStateRows\times\alphabet^\nSiteRows\rightarrow\mathcal{P}(\ancestralSplitSet).
$$
Then, given a site pattern--valued random variable $\alignmentColumnRV$, define the random variable
$$
\siteSplitRV := \patternToSplit(\alignmentColumnRV)
$$
that takes corresponding realizations $\siteSplit_j$ for some $j$, and
$$
\ancestralSplitRV := \ancestralToSplit(\alignmentColumnRV, \ancestralStateColumnRV)
$$
for a tip state--valued random variable $\alignmentColumnRV$ and an ancestral state--valued random variable $\ancestralStateColumnRV$.
\end{definition}
The mapping $\patternToSplit$ takes the complement of site patterns to obtain a site split in $\mathcal{P}(\siteSplitSet)$.
The mapping $\ancestralToSplit$ is defined by whether the tip states have their complements taken or not: if a set of tip labels $\alignmentColumn$ is in $\siteSplitSet$, $\ancestralToSplit(\alignmentColumn, \ancestralStateColumn)$ is $\ancestralStateColumn$; otherwise, if $\alignmentColumn$ is not in $\siteSplitSet$, then the complement of $\alignmentColumn$ necessarily is in $\siteSplitSet$, and $\ancestralToSplit(\alignmentColumn, \ancestralStateColumn)$ is the complement of $\ancestralStateColumn$.

For the $i$th factor of \eqref{eq:full_likelihood},
$$
\Pr(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t) = \Pr(\alignmentColumnRV=\alignmentColumn_i \mid \tau, t) \cdot \Pr(\ancestralStateColumnRV=\ancestralStateColumn_i \mid \alignmentColumnRV=\alignmentColumn_i, \tau, t).
$$
As a consequence of assuming a binary symmetric model, taking complements yields
\begin{align*}
    2\cdot \Pr(\alignmentColumnRV=\alignmentColumn_i \mid \tau, t) &= \Pr(\siteSplitRV=\patternToSplit(\alignmentColumn_i) \mid \tau, t) \\
                                                                 &= \Pr(\siteSplitRV=\siteSplit_j \mid \tau, t)
\end{align*}
for some $j$ and
\begin{align*}
    \Pr(\ancestralStateColumnRV=\ancestralStateColumn_i \mid \alignmentColumnRV=\alignmentColumn_i, \tau, t) &= \Pr(\ancestralSplitRV=\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i) \mid \siteSplitRV=\patternToSplit(\alignmentColumn_i), \tau, t).
\end{align*}
Given $(\tau, t)$, there exists an ordered list of sets $\fullAncestralSplitPartitions(\tau, t)=(\ancestralSplitPartition_1(\tau, t),\ldots,\ancestralSplitPartition_\nSiteSplits(\tau, t))$ such that any element $\xi_j$ of the $j$th component $\ancestralSplitPartition_j(\tau, t)$ satisfies
\begin{align*}
\max_{\ancestralSplit_k\in\mathcal{P}(\ancestralSplitSet)} \ \Pr(\ancestralSplitRV=\ancestralSplit_k \mid \siteSplitRV=\siteSplit_j, \tau, t) &= \Pr(\ancestralSplitRV = \xi_j \mid \siteSplitRV=\siteSplit_j, \tau, t).
\end{align*}
In other words, for the $j$th site split, $\ancestralSplitPartition_j(\tau, t)\subset\mathcal{P}(\ancestralSplitSet)$ is the set of most likely ancestral splits for that particular site split, topology and set of branch lengths, and $\xi_j$ is one of possibly many equiprobable ancestral state splits in $\ancestralSplitPartition_j(\tau, t)$.
For each $\alignmentColumn_i$, $\ancestralToSplit(\alignmentColumn_i, \cdot)$ is surjective, and from this we have
\begin{align*}
\max_{\ancestralStateColumn_i} \Pr(\ancestralSplitRV=\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i) \mid \siteSplitRV=\patternToSplit(\alignmentColumn_i), \tau, t) &= \Pr(\ancestralSplitRV = \xi_j \mid \siteSplitRV=\siteSplit_j, \tau, t).
\end{align*}

@@

Let $\xi_j$ be such a choice for each $1 \leq j \leq q$.
Then, the likelihood in \eqref{eq:profile_likelihood} written as a product over site patterns as opposed to sites is
\begin{align}
L_\nCols'(\tau, t; \fullAlignment) &= \max_{\fullAncestralStates} \ L_\nCols(\tau, t; \fullAlignment, \fullAncestralStates) \nonumber \\
                             &= \prod_{i=1}^{\nCols} \ \max_{\ancestralStateColumn_i} \ \Pr(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t) \nonumber \\
                             &\propto \prod_{i=1}^{\nCols} \ \max_{\ancestralStateColumn_i} \ \Pr(\siteSplitRV=\patternToSplit(\alignmentColumn_i) \mid \tau, t) \cdot \Pr(\ancestralSplitRV=\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i) \mid \siteSplitRV=\patternToSplit(\alignmentColumn_i), \tau, t) \nonumber \\
                             &= \prod_{i=1}^{\nCols} \ \Pr(\siteSplitRV=\patternToSplit(\alignmentColumn_i) \mid \tau, t) \cdot \max_{\ancestralStateColumn_i} \Pr(\ancestralSplitRV=\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i) \mid \siteSplitRV=\patternToSplit(\alignmentColumn_i), \tau, t) \nonumber \\
                             &= \prod_{j=1}^{\nSiteSplits} \ \left[\Pr(\siteSplitRV=\siteSplit_j \mid \tau, t)\cdot \Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j, \tau, t)\right] ^{\nCols_j(\fullAlignment)} \label{eq:site_pattern_likelihood}
\end{align}
where $\nCols_j(\fullAlignment)$ is the number of columns in $\fullAlignment$ that project to site split $\siteSplit_j$.

@@

We state the Hadamard representation of site split generating probabilities, following Section 8.6 of \citet{Semple2003-em}.
For each edge $e$ define the edge ``fidelity'' for that edge as
$$
\theta(e) = 1-2p(e).
$$
For an even-sized subset of $Y\subseteq\mathcal{S}$, we define the path set $P(Y)$ as the set of edges in the path connecting both elements of $Y$.
For $n$ taxa, the probability of observing site split $A\in\mathcal{P}(\siteSplitSet)$ is
\begin{equation}
\label{eq:hadamard_probability}
p_A = \frac{1}{2^{n-1}} \ \sum_{Y \subseteq \mathcal{S} : |Y| \equiv 0 (\mathrm{mod} \ 2)} \ \left[(-1)^{|Y \cap A|} \ \prod_{e\in P(Y)} \ \theta(e) \right].
\end{equation}
By convention, we set $P(\emptyset)=\emptyset$ and $\prod_{e\in\emptyset} \ \theta(e) = 1$.
For notational convenience, let
$$
p_{\siteSplit_j} := \Pr(\siteSplitRV=\siteSplit_j \mid \tau_1,t),
$$
for any site split $\siteSplit_j$.
Table~\ref{tab:sitepatprob} contains calculations of site pattern probabilities for our two topologies (Fig.~\ref{fig:farris-fels-top}).

>>>>

Since we have a finite character alphabet, for a given column $i$ there are a finite number of possible assignments of characters to tips $\alignmentColumn_i$ or internal nodes $\ancestralStateColumn_i$; this results in a simpler likelihood calculation.
Take the tip labels of $\tau$ to be $\{1,\ldots,\nSiteRows\}$.
For likelihood calculation under the binary symmetric model, we describe a given $\alignmentColumn_i$ as a subset of indices $\siteSplit\subseteq\siteSplitSet:=\{1,\ldots,\nSiteRows-1\}$ with equivalent characters, commonly called a ``site split.''
We define the site split $\siteSplit$ for a $\alignmentColumn_i$ as simply $\alignmentColumn_i$ if the label $\nSiteRows$ is not in $\alignmentColumn_i$, and as its complement otherwise.
Taking such a complement simplifies but does not change the result of likelihood computation because the probability of observing a particular collection of binary characters is equivalent to the probability of its complement under the binary symmetric model.

@@

We first fix notation.
\begin{definition}
Define the complement of $\alignmentColumn$ as $\overline{\alignmentColumn}$.
Let the mapping from site patterns to site splits
$$
\patternToSplit:\alphabet^\nSiteRows\rightarrow\mathcal{P}(\siteSplitSet)
$$
be
$$
\patternToSplit(\alignmentColumn) =
\left\{
    \begin{array}{ll}
        \alignmentColumn  & \mbox{if } \alignmentColumn\in\mathcal{P}(\siteSplitSet) \\
        \overline{\alignmentColumn}  & \mbox{if } \alignmentColumn\notin\mathcal{P}(\siteSplitSet)
    \end{array}
\right.
$$
and the mapping from ancestral states and tip states to ancestral state splits
$$
\ancestralToSplit:\alphabet^\nAncestralStateRows\times\alphabet^\nSiteRows\rightarrow\mathcal{P}(\ancestralSplitSet)
$$
be
$$
\ancestralToSplit(\alignmentColumn, \ancestralStateColumn) =
\left\{
    \begin{array}{ll}
        \ancestralStateColumn  & \mbox{if } \alignmentColumn\in\mathcal{P}(\siteSplitSet) \\
        \overline{\ancestralStateColumn}  & \mbox{if } \alignmentColumn\notin\mathcal{P}(\siteSplitSet).
    \end{array}
\right.
$$
Then, given a site pattern--valued random variable $\alignmentColumnRV$, define the random variable
$$
\siteSplitRV := \patternToSplit(\alignmentColumnRV)
$$
that takes corresponding realizations $\siteSplit_j$ for some $j$, and
$$
\ancestralSplitRV := \ancestralToSplit(\alignmentColumnRV, \ancestralStateColumnRV)
$$
for a tip state--valued random variable $\alignmentColumnRV$ and an ancestral state--valued random variable $\ancestralStateColumnRV$.
\end{definition}
The mapping $\patternToSplit$ operates by returning the complement of a site pattern to obtain a site split in $\mathcal{P}(\siteSplitSet)$ if it is not in $\mathcal{P}(\siteSplitSet)$.
The mapping $\ancestralToSplit$ is defined by whether the tip states have their complements taken or not: if a set of tip labels $\alignmentColumn$ is in $\siteSplitSet$, $\ancestralToSplit(\alignmentColumn, \ancestralStateColumn)$ is $\ancestralStateColumn$; otherwise, if $\alignmentColumn$ is not in $\siteSplitSet$, then the complement of $\alignmentColumn$ necessarily is in $\siteSplitSet$, and $\ancestralToSplit(\alignmentColumn, \ancestralStateColumn)$ is the complement of $\ancestralStateColumn$.

For the $i$th factor of \eqref{eq:full_likelihood},
$$
\Pr(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t) = \Pr(\alignmentColumnRV=\alignmentColumn_i \mid \tau, t) \cdot \Pr(\ancestralStateColumnRV=\ancestralStateColumn_i \mid \alignmentColumnRV=\alignmentColumn_i, \tau, t).
$$
As a consequence of assuming a binary symmetric model, for some $\siteSplit_j\in\mathcal{P}(\siteSplitSet)$ the mapping $\patternToSplit(\alignmentColumn_i)$ has the property
\begin{align*}
    \Pr(\siteSplitRV=\siteSplit_j \mid \tau, t) &= \Pr(\siteSplitRV=\patternToSplit(\alignmentColumn_i) \mid \tau, t)\\
    &= \Pr(\alignmentColumnRV=\alignmentColumn_i \cup \overline{\alignmentColumnRV}=\alignmentColumn_i \mid \tau, t)\\
    &= \Pr(\alignmentColumnRV=\alignmentColumn_i \mid \tau, t) + \Pr(\overline{\alignmentColumnRV}=\alignmentColumn_i \mid \tau, t)\\
    &= 2\cdot\Pr(\alignmentColumnRV=\alignmentColumn_i \mid \tau, t)
\end{align*}
where $\overline{Y}$ is the complement of the site-patterned random variable $Y$ and $\overline{Y}\sim Y$.
For ancestral states and for some $\ancestralSplit_k\in\mathcal{P}(\ancestralSplitSet)$
\begin{align*}
    \Pr(\ancestralStateColumnRV=\ancestralStateColumn_i \mid \alignmentColumnRV=\alignmentColumn_i, \tau, t) &= \Pr(\ancestralSplitRV=\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i) \mid \siteSplitRV=\patternToSplit(\alignmentColumn_i), \tau, t).
\end{align*}
Given $(\tau, t)$, there exist sets $\ancestralSplitPartition_1(\tau, t),\ldots,\ancestralSplitPartition_\nSiteSplits(\tau, t)$ such that $\xi_j\in\ancestralSplitPartition_j(\tau, t)$ satisfies
\begin{align*}
\max_{\ancestralSplit_k\in\mathcal{P}(\ancestralSplitSet)} \ \Pr(\ancestralSplitRV=\ancestralSplit_k \mid \siteSplitRV=\siteSplit_j, \tau, t) &= \Pr(\ancestralSplitRV = \xi_j \mid \siteSplitRV=\siteSplit_j, \tau, t).
\end{align*}
In other words, for the $j$th site split, $\ancestralSplitPartition_j(\tau, t)\subset\mathcal{P}(\ancestralSplitSet)$ is the set of most likely ancestral splits for that particular site split, topology and set of branch lengths, and $\xi_j$ is one of possibly many equiprobable ancestral state splits in $\ancestralSplitPartition_j(\tau, t)$.
For each $\alignmentColumn_i$, $\ancestralToSplit(\alignmentColumn_i, \cdot)$ is surjective as it maps all possible values from $\alphabet^\nAncestralStateRows$ to an element in $\mathcal{P}(\ancestralSplitSet)$, and from this we have
\begin{align*}
    \max_{\ancestralStateColumn_i} \ \Pr(\ancestralStateColumnRV=\ancestralStateColumn_i \mid \alignmentColumnRV=\alignmentColumn_i, \tau, t) &= \max_{\ancestralStateColumn_i} \ \Pr(\ancestralSplitRV=\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i) \mid \siteSplitRV=\patternToSplit(\alignmentColumn_i), \tau, t)\\
    &= \max_{\ancestralSplit_k\in\mathcal{P}(\ancestralSplitSet)} \ \Pr(\ancestralSplitRV=\ancestralSplit_k \mid \siteSplitRV=\siteSplit_j, \tau, t)\\
    &= \Pr(\ancestralSplitRV = \xi_j \mid \siteSplitRV=\siteSplit_j, \tau, t).
\end{align*}
Thus, each term in the likelihood can be collapsed into terms relating only to site splits, indexed by $j$,  as opposed to each individual observation, indexed by $i$.

@@

The likelihood in \eqref{eq:profile_likelihood} written as a product over site patterns as opposed to sites is
\begin{align}
L_\nCols'(\tau, t; \fullAlignment) &= \max_{\fullAncestralStates} \ L_\nCols(\tau, t; \fullAlignment, \fullAncestralStates) \nonumber \\
                             &= \prod_{i=1}^{\nCols} \ \max_{\ancestralStateColumn_i} \ \Pr(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t) \nonumber \\
                             &\propto \prod_{i=1}^{\nCols} \ \max_{\ancestralStateColumn_i} \ \Pr(\siteSplitRV=\patternToSplit(\alignmentColumn_i) \mid \tau, t) \cdot \Pr(\ancestralSplitRV=\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i) \mid \siteSplitRV=\patternToSplit(\alignmentColumn_i), \tau, t) \nonumber \\
                             &= \prod_{i=1}^{\nCols} \ \Pr(\siteSplitRV=\patternToSplit(\alignmentColumn_i) \mid \tau, t) \cdot \max_{\ancestralStateColumn_i} \Pr(\ancestralSplitRV=\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i) \mid \siteSplitRV=\patternToSplit(\alignmentColumn_i), \tau, t) \nonumber \\
                             &= \prod_{j=1}^{\nSiteSplits} \ \left[\Pr(\siteSplitRV=\siteSplit_j \mid \tau, t)\cdot \Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j, \tau, t)\right] ^{\nCols_j(\fullAlignment)} \label{eq:site_pattern_likelihood}
\end{align}
for some $\xi_j\in\ancestralSplitPartition_j(\tau, t)$ with $1 \le j \le q$ where $\nCols_j(\fullAlignment)$ is the number of columns in $\fullAlignment$ that project to site split $\siteSplit_j$.

@@

We state the Hadamard representation of site split generating probabilities---that is, probabilities of obtaining particular site splits given a tree---following Section 8.6 of \citet{Semple2003-em}.
For each edge $e$ define the edge ``fidelity'' for that edge as
$$
\theta(e) = 1-2p(e)
$$
where $p(e)$ is the probability of a character change along edge $e$.
For an even-sized subset of $Y\subseteq\mathcal{S}$, we define the path set $P(Y)$ as the set of edges in the path connecting both elements of $Y$.
For $n$ taxa, the probability of observing site split $A\in\mathcal{P}(\siteSplitSet)$ is
\begin{equation}
\label{eq:hadamard_probability}
p_A = \frac{1}{2^{n-1}} \ \sum_{Y \subseteq \mathcal{S} : |Y| \equiv 0 (\mathrm{mod} \ 2)} \ \left[(-1)^{|Y \cap A|} \ \prod_{e\in P(Y)} \ \theta(e) \right].
\end{equation}
By convention, we set $P(\emptyset)=\emptyset$ and $\prod_{e\in\emptyset} \ \theta(e) = 1$.
For notational convenience, let
$$
p_{\siteSplit_j} := \Pr(\siteSplitRV=\siteSplit_j \mid \tau_1,t),
$$
for any site split $\siteSplit_j$.
Table~\ref{tab:sitepatprob} contains calculations of site pattern probabilities for our two topologies (Fig.~\ref{fig:farris-fels-top}).
